% ============================================================
%  ppo_queueing_report.tex
%  Technical note: PPO for the deadline-constrained multi-server
%  queue --- the method, the stabilization recipe, and benchmark
%  results on Experiments 2 and 3.
%
%  Intended reader: comfortable with MDPs, uniformisation and
%  RVI; no prior exposure to policy-gradient RL assumed.
% ============================================================
\documentclass[11pt,a4paper]{article}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{parskip}

\newtheorem{remark}{Remark}
\newtheorem{mechanism}{Mechanism}

\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\clip}{\operatorname{clip}}

\title{Proximal Policy Optimization for a Deadline-Constrained
       Multi-Server Queue:\\
       Method, Stabilization Recipe, and Benchmark Results}
\author{Technical Note}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We describe how Proximal Policy Optimization (PPO) was made to learn
(near-)optimal scheduling policies for our deadline-constrained
multi-server queueing MDP, benchmarked against exact Relative Value
Iteration (RVI) solutions.
Section~\ref{sec:ppo} introduces PPO from first principles for a reader
familiar with MDPs and RVI but not with policy-gradient reinforcement
learning.
Section~\ref{sec:ours} presents our specific formulation: the semi-MDP
discounting correction, potential-based reward shaping, value-target
normalization, staggered environment resets, and guarded temperature
annealing --- each introduced as the fix to a specific, diagnosed failure
mechanism.
Section~\ref{sec:experiments} defines the two benchmark instances
(Experiment~3: routing with a specialist and a generalist server;
Experiment~2: fully flexible servers with asymmetric costs, where the
optimal policy requires \emph{strategic idling}) and reports results:
Experiment~3 is solved (7--8/8 seeds within 5\% of RVI), Experiment~2 is
solved per problem instance (7/8 seeds within 6\%; failures are a seed
lottery mitigated by training 2--3 seeds and selecting by evaluation).
A controlled single-ingredient ablation study (8 cells $\times$ 8 seeds)
substantiates each ingredient of the recipe: removing any one of the four
main stabilizers collapses the success rate from 6--7/8 to 0--3/8, with
failures landing on two exactly reproducible attractors (never-serve and
FIFO-clone).
Section~\ref{sec:ql} extends the study to the \emph{queue-lateness}
objective, where the failure landscape inverts: the extraction gap
disappears but heavy-tailed returns defeat both the policy gradient and
the annealing guard, and the binary-cost recipe converges to a FIFO clone
on 8/8 seeds.  A different recipe --- the average-reward objective at
$64\times$ batch size with a tail-robust guard --- restores near-optimal
learning (7/8 seeds beat FIFO; best seeds at the RVI optimum), after the
RVI benchmark itself is repaired (learned policies exposed a 6\%
truncation bias).
Finally, Section~\ref{sec:pe} tests the diagnosis constructively: a
\emph{per-event action space} (each free server capacity unit picks a
job type or idles, one credit-concentrated decision each) eliminates the
mechanisms rather than patching them.  In the reformulated space, PPO
\emph{without} temperature annealing reaches the exact RVI optimum on
8/8 seeds --- the seed lottery disappears, and the extraction scaffolding
turns from load-bearing to harmful --- while DCL escapes its FIFO-clone
fixed point (6/8 optimal) and loses its queue-lateness collapses (8/8
beat FIFO).
\end{abstract}

\tableofcontents
\newpage

% ============================================================
\section{Setting (recapitulation)}
\label{sec:setting}
% ============================================================

We briefly restate the system and its MDP; full details are in the
companion RVI note.

\paragraph{System.}
$K$ server pools serve $N$ job types.  Jobs of type $n$ arrive as
independent Poisson streams with rate $\lambda_n$ and wait in per-type
FIFO queues.  Service is exponential; server pool $k$ can serve a subset
of types at rates $\mu_{k,n}$.  Each type has a physical deadline
$T_n$ and a cost rate $c_n$: a waiting job that exceeds its deadline
incurs cost.  The objective is the long-run average cost.

\paragraph{Uniformised MDP.}
The continuous-time dynamics are embedded into a discrete-time chain by
uniformisation.  At each \emph{period} exactly one event occurs:
an \textbf{arrival} of type $n$ (probability $\propto \lambda_n$),
a \textbf{service completion} (probability $\propto$ busy rates),
a \textbf{tick} that advances all waiting-time clocks by one unit
(probability $\propto \nu$, the tick rate\footnote{The RVI note denotes
the tick rate by $\gamma$; here we reserve $\gamma$ for the discount
factor and write $\nu$ for the tick rate.}), or
\textbf{nothing} (self-loop).  The uniformisation rate is
$\Lambda = \sum_n \lambda_n + \nu + \sum \text{(busy service rates)}$.
The state is $s = (f_0,\dots,f_{N-1},\, B,\, \alpha,\, \kappa)$: per-type
first-in-line (FIL) waiting times $f_n$ (in ticks; $-1$ if empty), the
busy configuration $B$, the action counter $\alpha$, and the category
$\kappa \in \{\textsc{AwaitEvent}, \textsc{AwaitAction}\}$.  Deadlines in
tick units are $d_n = T_n \nu$.

\paragraph{Action queue.}
Whenever idle capacity and waiting jobs coexist (after each tick), the
feasible (server, job-type) assignment candidates are collected into an
ordered \emph{action queue} (descending FIL, i.e.\ FIFO order) and
presented \emph{sequentially}.  At each position the agent chooses
action~$1$ (\emph{assign} the presented candidate) or action~$0$
(\emph{skip}, advance to the next candidate).  Skipping the final
candidate means the system idles until the next event.  Several decisions
may therefore occur within a single period, and a skipped job is
re-presented at the next tick.

\paragraph{Cost.}
The ground-truth objective (\texttt{reward\_type}~0) is the binary
deadline cost, charged at \textsc{AwaitEvent} states:
\begin{equation}
  c_{\mathrm{bin}}(s) = \sum_n c_n \,\mathbf{1}\{f_n > d_n\}.
  \label{eq:bincost}
\end{equation}
Policies are evaluated by simulation (100 trajectories $\times$
$5\cdot10^5$ periods) and reported as cost per unit time, i.e.\ mean cost
per period times $\Lambda$; we write this as
$\mathrm{cost}\times\Lambda$.  RVI supplies the exact optimal benchmark
$g^*$ on both instances, and the greedy FIFO policy (always assign the
oldest feasible candidate) is the natural heuristic baseline.

\begin{remark}[Choice of objective]
All experiments in this note use the \emph{binary} deadline cost
\eqref{eq:bincost}, whereas parts of our earlier RVI analysis used the
\emph{queue-lateness} cost (linear excess over the deadline, summed over
the queue).  The two are different objectives with values on different
scales; all RVI and FIFO reference values below are binary-cost values.
The binary cost is the \emph{harder} learning target: as the RVI
heatmap analysis showed, its near-zero gradient away from the deadline
creates action-value near-ties across most of the state space, and
these near-ties are precisely what makes policy-gradient learning and
argmax extraction difficult (Sections~\ref{sec:shaping}
and~\ref{sec:forgiveness}).  The reward shaping introduced below can be
read as restoring, in a cost-neutral way, the pre-deadline signal
density that the queue-lateness objective has natively.
\end{remark}

% ============================================================
\section{PPO in general terms}
\label{sec:ppo}
% ============================================================

\subsection{From value iteration to policy search}

RVI computes the exact optimal policy by sweeping the entire state space.
This is infeasible beyond small instances, so we turn to methods that
(i)~replace enumeration by \emph{simulation}, and (ii)~replace tables by
\emph{function approximation}.  Policy-gradient methods do both directly
in policy space: fix a parametric family of \emph{stochastic} policies
$\pi_\theta(a \mid s)$ --- here a neural network mapping a feature vector
of $s$ to a probability distribution over actions --- and improve
$\theta$ by stochastic gradient ascent on the expected return.

We use the discounted objective as a proxy for the average-cost
objective:
\begin{equation}
  J(\theta) = \E_{\pi_\theta}\Bigl[\sum_{t\ge 0} \gamma^{t}\, r_t\Bigr],
\end{equation}
with $\gamma$ close to $1$.  The effective horizon is
$1/(1-\gamma)$ periods, i.e.\ $1/\bigl((1-\gamma)\Lambda\bigr)$ time
units; with our $\gamma = 0.997$ this is ${\sim}330$ periods, long enough
to span many deadline windows.  (Rewards are negative costs; maximizing
return minimizes cost.)

\subsection{The policy gradient and the advantage function}

The policy gradient theorem states
\begin{equation}
  \nabla_\theta J(\theta)
  = \E_{\pi_\theta}\bigl[\nabla_\theta \log \pi_\theta(a_t \mid s_t)\,
    A^{\pi_\theta}(s_t, a_t)\bigr],
  \qquad
  A^{\pi}(s,a) = Q^{\pi}(s,a) - V^{\pi}(s).
  \label{eq:pg}
\end{equation}
The interpretation is direct: increase the log-probability of actions
that performed \emph{better than the policy's average} in their state
(positive advantage), decrease it otherwise.  Two consequences of
\eqref{eq:pg} matter throughout this note:
\begin{itemize}
  \item The update is \emph{advantage-weighted}: in states where all
        actions perform nearly equally under the current policy
        ($A \approx 0$), the gradient is negligible --- \emph{such states
        are never explicitly trained}, however frequently they are
        visited.
  \item The expectation is under the current policy's own state
        distribution: the data the algorithm learns from is generated by
        the very policy being trained, so a degraded policy also degrades
        its own future training data.
\end{itemize}

The advantage is unknown and must be estimated.  A \emph{critic}
$V_\phi(s)$ (a second network output) is regressed onto observed
returns, and the advantage is estimated by Generalized Advantage
Estimation (GAE): with the temporal-difference error
$\delta_t = r_t + \gamma V_\phi(s_{t+1}) - V_\phi(s_t)$,
\begin{equation}
  \hat A_t \;=\; \sum_{l \ge 0} (\gamma \lambda)^{l}\, \delta_{t+l},
  \label{eq:gae}
\end{equation}
where $\lambda \in [0,1]$ trades off bias ($\lambda = 0$: pure one-step
TD, biased by critic error) against variance ($\lambda = 1$: pure Monte
Carlo).\footnote{Throughout, plain $\lambda$ is the GAE parameter;
arrival rates always carry a subscript ($\lambda_n$).}

\subsection{The PPO clipped update}

Vanilla policy gradient permits exactly one gradient step per batch of
simulated experience: after the step, the data is off-policy and
\eqref{eq:pg} no longer holds.  PPO makes the data reusable for several
epochs of minibatch updates by controlling how far the new policy may
move from the policy $\pi_{\theta_{\mathrm{old}}}$ that generated the
data.  With the likelihood ratio
$\rho_t(\theta) = \pi_\theta(a_t \mid s_t)\,/\,
\pi_{\theta_{\mathrm{old}}}(a_t \mid s_t)$,
PPO maximizes the \emph{clipped surrogate}
\begin{equation}
  L^{\mathrm{clip}}(\theta) =
  \E_t\Bigl[\min\bigl(\rho_t \hat A_t,\;
      \clip(\rho_t,\, 1-\epsilon,\, 1+\epsilon)\, \hat A_t\bigr)\Bigr],
  \label{eq:clip}
\end{equation}
with $\epsilon = 0.2$.  The clip removes the incentive to push
$\rho_t$ beyond $1 \pm \epsilon$: once an action's probability has moved
by that factor, further movement earns no additional surrogate reward.
This is a cheap, first-order stand-in for a trust region.

The full loss combines three terms:
\begin{equation}
  L(\theta,\phi) \;=\; -L^{\mathrm{clip}}(\theta)
  \;+\; c_v \,\E_t\bigl[(V_\phi(s_t) - \hat R_t)^2\bigr]
  \;-\; c_e \,\E_t\bigl[\mathcal H\bigl(\pi_\theta(\cdot \mid s_t)\bigr)\bigr],
  \label{eq:loss}
\end{equation}
where $\hat R_t = \hat A_t + V_\phi(s_t)$ is the return target,
$\mathcal H$ is the policy entropy (a bonus that sustains exploration by
penalizing premature determinism), and $c_v, c_e$ are weights.  Both
heads share one network trunk, so the two regression problems interact
--- a fact that becomes important in Section~\ref{sec:vnorm}.

One training \emph{update} is: simulate $E$ parallel environments for
$T$ decisions each under the current policy; compute
\eqref{eq:gae} on the batch of $E \cdot T$ transitions; then take
several epochs of minibatch gradient steps on \eqref{eq:loss}.

\subsection{From stochastic policy to deterministic decision rule}
\label{sec:extraction}

PPO trains a stochastic policy, but what we ultimately want is a
deterministic scheduling rule.  The standard extraction is the
\emph{argmax policy}: in each state, take the action with the highest
probability.  It is worth flagging already here that this step is
\emph{not} innocuous: training only guarantees that the stochastic
policy performs well, and the argmax can differ from the stochastic
policy precisely in the states where the learned probabilities are close
to $1/2$.  Section~\ref{sec:forgiveness} shows that in our MDP this gap
is systematic rather than incidental, and
Section~\ref{sec:ablation} exhibits it empirically.

% ============================================================
\section{Our PPO formulation and the stabilization recipe}
\label{sec:ours}
% ============================================================

Applied naively, PPO fails on our MDP: on Experiment~2 it collapsed to
the ``never serve'' policy on 8 of 8 seeds ($9.9\times$ the optimal
cost).  Every ingredient below is the fix for a specific, diagnosed
failure mechanism; the ablation study in Section~\ref{sec:ablation}
substantiates each one in a controlled way.  We describe the
formulation mathematically; the concrete hyperparameters are collected
in Table~\ref{tab:hyper}.

\subsection{Semi-MDP timing: discounting by elapsed periods}
\label{sec:dper}

Decisions occur on the action queue, so consecutive decisions may be
separated by \emph{zero} periods (two candidates presented within the
same tick) or by \emph{many} periods (idle stretches between ticks).
The MDP between decisions is a semi-Markov process, and the discount
between decision $t$ and decision $t{+}1$ must be
\begin{equation}
  \gamma^{\Delta_t}, \qquad
  \Delta_t = \#\{\text{periods elapsed between the two decisions}\}
  \in \{0, 1, 2, \dots\},
\end{equation}
used both in the TD errors
$\delta_t = r_t + \gamma^{\Delta_t} V_\phi(s_{t+1}) - V_\phi(s_t)$
and in the GAE recursion (decay $\gamma^{\Delta_t}\lambda$).

\begin{mechanism}[Idleness subsidy through mis-discounting]
\label{mech:dper}
An earlier implementation discounted \emph{every} decision by one period
($\gamma^{\max(1,\Delta_t)}$, effectively).  An intra-tick decision
chain then discounts the future as if time passed at every skip, so a
policy that skips through the whole queue reaches the next (costly)
event with extra discount applied --- idleness is \emph{subsidized}, and
most strongly in busy states where the action queue is long.  This
single error was sufficient to push training into the never-serve
attractor.  The fix is the correct semi-MDP exponent with
$\gamma^0 = 1$.
\end{mechanism}

\subsection{Potential-based reward shaping (the urgency ramp)}
\label{sec:shaping}

The binary deadline cost \eqref{eq:bincost} is nearly flat: a job
incurs no cost until its deadline passes.  During early training almost
all sampled decisions therefore have $A \approx 0$, while the rare
deadline penalty arrives many decisions after the skip that caused it.
Against this sparse signal stands a strong attractor: never serving
makes the immediate future of every individual decision look identical,
and by Mechanism~\ref{mech:dper}-type credit-assignment noise the policy
drifts there easily.

We therefore shape the reward with a potential function.  Let
$\Phi(s)$ be the total \emph{accumulated urgency} of the jobs currently
waiting: each tick, the FIL job of type $n$ adds
\begin{equation}
  c_n \,\frac{\min(f_n, d_n)}{d_n}
  \label{eq:ramp}
\end{equation}
to its accumulated urgency (a ramp that rises linearly toward the
deadline, then saturates), and a job's accumulated urgency is
\emph{refunded} in full at the moment it enters service.  The shaped
per-transition cost is the base cost plus the accrual minus the refund,
i.e.\ exactly the potential-based form
$r' = r + \Phi(s) - \Phi(s')$ (in cost convention).  Potential-based
shaping leaves the optimal policy unchanged and, for stationary policies
under the average-cost criterion, leaves the long-run average unchanged
(the potential term telescopes).  We verified this empirically: FIFO and
RVI evaluate identically to four decimal places with and without
shaping, so RVI remains a valid benchmark for shaped training runs.

The effect on learning is dense, immediate credit: every tick spent not
serving an urgent job is charged \emph{now}, and serving refunds
\emph{now}.  Shaping is load-bearing: without it the success rate drops
from $6$--$7/8$ seeds to $2/8$ (Section~\ref{sec:ablation}).

\subsection{Value-target normalization}
\label{sec:vnorm}

Returns in this MDP are large (orders of $10^2$--$10^3$ in cost units),
while the policy-gradient part of \eqref{eq:loss} operates on
normalized advantages of order~$1$.

\begin{mechanism}[Value-scale domination]
With raw return targets, the value MSE term in \eqref{eq:loss} is orders
of magnitude larger than the policy term, and its gradients monopolize
the shared trunk: the representation is shaped for value regression
while the policy head starves.
\end{mechanism}

The fix: maintain a running estimate $\sigma_R$ of the return standard
deviation, train the value head on $\hat R_t / \sigma_R$, and rescale
its output by $\sigma_R$ wherever a raw value is needed (bootstrap
targets, GAE).  Advantages are additionally standardized per batch.
Without this ingredient, \emph{zero} of eight seeds succeed: every run
converges to one of the two attractor policies (never-serve or a FIFO
clone; Section~\ref{sec:ablation}).

\subsection{Persistent environments and staggered resets}
\label{sec:resets}

The $E = 16$ parallel simulation environments are \emph{persistent}:
the infinite-horizon chain continues across updates, which matches the
average-cost objective.  It also creates a trap.

\begin{mechanism}[Data poisoning by deep-queue excursions]
\label{mech:poison}
One bad excursion (e.g.\ a temporarily idle-happy policy) drives an
environment into a deep-backlog region where the shaping ramps
\eqref{eq:ramp} are saturated and no dense gradient exists.  If all
environments drift there, the training data no longer contains the
healthy short-queue states at all; the policy's behavior on those
states then degrades unmonitored (weights move on shared features), and
evaluation --- which starts from the empty state --- collapses.  This is
self-reinforcing through the on-policy data distribution and is not
recoverable by any gradient-side guard: only fresh-start data helps.
\end{mechanism}

The fix is \emph{staggered resets}: each update, one environment (round
robin, period $16$ updates) is re-initialized to the empty state.  The
data distribution stays anchored to the region the deployed policy
starts in, at a negligible cost in on-policyness.  Ablation: with no
resets, $1/8$ seeds succeed and five of eight converge to the exact
never-serve policy.

\subsection{Exploration and extraction: entropy and guarded
temperature annealing}
\label{sec:anneal}

\subsubsection{The re-presentation forgiveness mechanism}
\label{sec:forgiveness}

Recall from Section~\ref{sec:extraction} that the deployed policy is the
argmax of the trained stochastic policy.  In this MDP the gap between
the two is systematic.  A job skipped now is re-presented at every
subsequent tick while it waits.  Suppose the policy assigns probability
$p$ to serving a given job, and the job will be re-presented $m$ more
times within its deadline slack.  The probability it is served in time
is
\begin{equation}
  1 - (1-p)^{m+1},
\end{equation}
which for moderate $p$ and a few ticks of slack is already close to
$1$: with $p = 0.4$ and $m = 8$, the job is served in time with
probability ${\approx}0.99$.  \emph{The stochastic policy is
behaviorally near-optimal in such states even when $p$ is on the wrong
side of $\tfrac12$} --- so the advantage signal vanishes
(cf.\ the advantage-weighting in \eqref{eq:pg}) and training never
corrects the sign.  The argmax, however, turns $p = 0.4$ into
\emph{never serve this job here}, deterministically, at every
re-presentation.  Wrong-sign near-ties are invisible to training and
fatal to extraction.

\subsubsection{Guarded behavior-temperature annealing}

The mitigation is to make the \emph{behavior} policy converge toward the
argmax policy during training, so that states where the argmax action is
wrong actually hurt the observed return and receive gradient.  We sample
rollout actions from a tempered policy
\begin{equation}
  \pi^{(T)}_\theta(a \mid s) \;\propto\;
  \exp\bigl(z_\theta(s,a)/T\bigr),
\end{equation}
where $z_\theta$ are the policy logits and the temperature $T$ anneals
from $1$ toward $T_{\min}$ during the second half of training.
Annealing \emph{blindly} on a clock is dangerous --- it amplifies
whichever mode the policy is currently in, sharpening bad seeds into
collapse --- so the schedule is \emph{guarded}: $T$ steps down (factor
$0.9$) only while an exponential moving average of the rollout reward
rate is not degrading relative to its best level since annealing began,
and steps back up when it is.  The health signal is measured
\emph{per period}, never per decision: an idling policy makes many
zero-reward intra-tick decisions, so a per-decision mean dilutes
precisely when the policy degrades toward idleness.

Two auxiliary devices complete the scheme.  A \emph{best-sharp snapshot}
stores the network parameters whenever behavior is both sharp
($T \le 2T_{\min}$) and healthy, and is restored at the end of training
unless the final parameters match it --- the guard's late retreats
otherwise discard the best policy of the run.  At evaluation time we
extract a small \emph{readout battery} (argmax; tempered stochastic;
argmax with a bias added to the serve logit) and select by evaluation;
on hard instances this converts a per-seed success rate into a
per-problem one.  Separately, the entropy coefficient $c_e$ in
\eqref{eq:loss} is annealed linearly to zero over the second half of
training, so the policy is allowed to become deterministic exactly when
the temperature schedule wants it to.

\subsection{What remains: advantage-weighted blindness}
\label{sec:blindness}

The mechanisms above remove the systematic failures; what remains on the
hardest instance is a residual \emph{seed lottery}
(Section~\ref{sec:exp2}).  Its root cause is the advantage-weighting in
\eqref{eq:pg} combined with Section~\ref{sec:forgiveness}: some
decision boundaries are behaviorally flat under the stochastic policy
and never receive gradient, and whether the argmax lands on the right
side of them is decided by initialization noise.  More compute helps
some seeds (doubling to $30{,}000$ updates rescues several
$6{,}000$-update failures); an auxiliary advantage head (a third network
output regressing the observed GAE advantages, giving an alternative
readout) rescued individual seeds but showed no systematic improvement
in an 8-seed A/B test ($5/8$ vs $5/8$).  The principled fix --- an
action-space reformulation in which each freed server triggers exactly
one decision (``which class to serve, or idle'') so that skipping is a
real commitment rather than a re-presentable preference --- is known but
deliberately deferred; a first probe in that direction is reported in
Section~\ref{sec:skipall}.

\subsection{Hyperparameters}

\begin{table}[h]
\centering
\caption{The tuned recipe (all experiments in
Section~\ref{sec:experiments} unless stated otherwise).}
\label{tab:hyper}
\begin{tabular}{llll}
\toprule
discount $\gamma$ & $0.997$ &
environments $E$ & $16$ (persistent) \\
GAE $\lambda$ & $0.95$ &
rollout $T$ & $512$ decisions \\
clip $\epsilon$ & $0.2$ &
updates & $6000$ \\
entropy coef.\ $c_e$ & $0.03$, annealed $\to 0$ &
epochs $\times$ minibatch & $4 \times 256$ \\
value coef.\ $c_v$ & $0.5$ &
learning rate & $3\cdot10^{-4}$ (Adam) \\
temperature floor $T_{\min}$ & $0.1$, guarded &
reset period & $16$ updates \\
network & \multicolumn{3}{l}{MLP $64 \times 32$, shared trunk; policy,
value (and aux-advantage) heads} \\
\bottomrule
\end{tabular}
\end{table}

Two calibration notes.  First, $\gamma = 0.997$ is load-bearing:
$0.99$ plateaus at a FIFO clone (effective horizon too short to see the
benefit of strategic idling), $0.999$ destabilizes credit assignment;
the sensitivity sweep (three seeds per knob) puts both alternatives at
$3$--$5\times$ higher mean cost ratios.  Second, the variance knobs
matter: raising the entropy coefficient from $0.01$ to $0.03$ and
lowering $T_{\min}$ from $0.25$ to $0.1$ each reduced seed-to-seed
standard deviation by a factor ${\sim}4$ at equal mean.

% ============================================================
\section{Experiments and results}
\label{sec:experiments}
% ============================================================

\subsection{The two benchmark instances}

Both instances use tick rate $\nu = 3$ and the binary deadline cost
\eqref{eq:bincost} as ground truth (training uses its shaped version,
Section~\ref{sec:shaping}).  Reported numbers are cost per unit time
($\mathrm{cost}\times\Lambda$); lower is better.

\paragraph{Experiment 3 (routing: specialist + generalist).}
$N = 2$ types, $\lambda = (0.8, 0.2)$, costs $c = (100, 300)$, deadline
$T_n = 1$; server pool $0$ is a \emph{specialist} (type $0$ only,
$\mu = 1$), pool $1$ a \emph{generalist} (both types, $\mu = 1$).  The
optimal policy must protect the generalist's capacity for type-$1$ jobs.
Benchmarks: $\mathrm{RVI}\times\Lambda = 18.96$,
$\mathrm{FIFO}\times\Lambda = 24.15$ (ratio $1.27$).

\paragraph{Experiment 2 (strategic idling: fully flexible, asymmetric
cost).}
$N = 2$ types, $\lambda = (0.25, 0.25)$, costs $c = (100, 300)$,
deadlines $T_n = 3$; two fully flexible single-server pools with
$\mu = 0.35$ ($\rho \approx 0.71$).  Since every server can serve every
job, the \emph{only} way to beat FIFO is strategic idling: refusing a
cheap type-$0$ job now to keep capacity free for a possible expensive
type-$1$ arrival.  Benchmarks: $\mathrm{RVI}\times\Lambda = 40.50$,
$\mathrm{FIFO}\times\Lambda = 70.46$ (ratio $1.74$).  This is the harder
instance: the payoff of idling is realized only through rare future
events, exactly the signal that Sections~\ref{sec:forgiveness}
and~\ref{sec:blindness} identify as hardest to learn.

``Within 5\%'' below means argmax cost ratio to RVI at most $1.05$.

\subsection{Experiment 3: solved}
\label{sec:exp3}

With the recipe of Section~\ref{sec:ours} (at the pre-tuning variance
settings $c_e = 0.01$, $T_{\min} = 0.25$, rollout $256$, plus the
auxiliary head), the eight seeds give
\[
  \mathrm{argmax}\times\Lambda =
  19.12,\; 19.88,\; 19.16,\; 18.97,\; 19.37,\; 18.95,\; 19.09,\; 21.55
\]
against $\mathrm{RVI}\times\Lambda = 18.96$: \textbf{7/8 within 5\%}
(the eighth at $13.7\%$), all far below FIFO's $24.15$.  Training is
${\sim}10$ minutes per seed on a shared 16-core node slice.  We consider
this instance solved and use it below mainly as the easy control.

\subsection{Experiment 2: solved per problem, seed lottery per run}
\label{sec:exp2}

Table~\ref{tab:exp2} shows the argmax results across eight seeds for
three budgets of the original recipe, and for the variance-tuned recipe
(Table~\ref{tab:hyper}).

\begin{table}[h]
\centering
\caption{Experiment 2, argmax $\mathrm{cost}\times\Lambda$ vs
$\mathrm{RVI}\times\Lambda = 40.50$ (FIFO $= 70.46$).  $\times$ denotes
collapse ($> 2\times$ RVI).  ``+aux'' adds the auxiliary advantage head;
``tuned'' is Table~\ref{tab:hyper}.}
\label{tab:exp2}
\begin{tabular}{lcccc}
\toprule
Seed & $6000$u & $6000$u+aux & $30\,000$u & tuned, $6000$u \\
\midrule
1 & 44.78 & 40.80 & 40.85 & 42.84 \\
2 & $\times$ & 41.99 & 41.73 & 42.20 \\
3 & 44.81 & 53.27 & 60.84 & 40.78 \\
4 & $\times$ & 42.00 & 41.58 & 42.51 \\
5 & 40.81 & $\times$ & 40.78 & 41.02 \\
6 & 40.92 & 41.59 & 42.37 & 40.78 \\
7 & 40.90 & 40.89 & 41.89 & 40.93 \\
8 & $\times$ & 57.48 & $\times$ & $\times$ \\
\midrule
within 5\% & 4/8 & 5/8 & 6/8 & 7/8\footnotemark \\
\bottomrule
\end{tabular}
\end{table}
\footnotetext{Seeds 1 and 4 sit at $5.8\%$ and $5.0\%$; counting
strictly, 5/8 within 5\% and 7/8 within 6\%.}

Three observations.  First, the tuned recipe is both better and much
tighter: seven of eight seeds land in $[40.78,\, 42.84]$.  Second,
failures do not attach to specific seeds: seed 3 fails with the
auxiliary head and succeeds without; seed 5 the reverse; only seed 8
fails in most (not all) configurations.  This is a lottery over
initializations, not a set of hard seeds --- consistent with
Section~\ref{sec:blindness}.  Third, and practically decisive: training
2--3 seeds and selecting by evaluation yields a within-5\% policy
\emph{every} time; per problem instance, Experiment~2 is solved at a
cost of ${\sim}30$ minutes of compute.  The runs are exactly
reproducible (same seed, same result to four decimals), so the tables
above are deterministic artifacts, not noisy samples.

\subsection{Controlled ablations: substantiating each ingredient}
\label{sec:ablation}

The mechanisms of Section~\ref{sec:ours} were diagnosed during
development, i.e.\ on a moving codebase.  To substantiate them, we
re-ran everything as controlled single-ingredient ablations from the
frozen tuned recipe: eight cells $\times$ eight seeds, identical in
every respect except the named ingredient
(Table~\ref{tab:ablation}).

\begin{table}[h]
\centering
\caption{Ablation matrix on Experiment 2: argmax cost ratio to RVI,
8 seeds per cell.  Every cell is the tuned recipe with exactly one
change.  For reference: never-serve evaluates at ratio $9.876$, FIFO at
$1.738$.}
\label{tab:ablation}
\begin{tabular}{llccc}
\toprule
Cell & Removes / restores & within 5\% & median & worst \\
\midrule
\texttt{base} & (control) & 6/8 & 1.03 & 5.41 \\
\texttt{dper} & restores mis-discounting (M.\ref{mech:dper}) & 0/8 & 7.26 & 9.85 \\
\texttt{vnorm} & value normalization off (\S\ref{sec:vnorm}) & 0/8 & 9.88 & 9.88 \\
\texttt{resets0} & no environment resets (\S\ref{sec:resets}) & 1/8 & 9.88 & 9.88 \\
\texttt{shape0} & no reward shaping (\S\ref{sec:shaping}) & 2/8 & 1.87 & 9.69 \\
\texttt{anneal0} & no temperature annealing (\S\ref{sec:anneal}) & 3/8 & 1.29 & 7.75 \\
\bottomrule
\end{tabular}
\end{table}

The matrix confirms the recipe ingredient by ingredient, and the
\emph{failure signatures} carry as much information as the counts:

\begin{itemize}
  \item \textbf{Mis-discounting} (\texttt{dper}): total failure, with
  most seeds at ratios $8.9$--$9.85$, i.e.\ at or near the never-serve
  attractor --- the direction Mechanism~\ref{mech:dper} predicts
  (idleness subsidy), not merely ``worse''.
  \item \textbf{Value scale} (\texttt{vnorm}): perfectly bimodal.  Every
  seed converges to \emph{exactly} ratio $9.876$ (never-serve; the value
  matches an independently simulated always-skip policy to four
  decimals) or \emph{exactly} $1.738$ (a FIFO clone, matching simulated
  FIFO to four decimals).  With a crushed trunk the policy can only fall
  into one of the two attractors.
  \item \textbf{Resets} (\texttt{resets0}): seven of eight seeds
  collapse (five to exact never-serve); one lucky seed solves the
  instance --- Mechanism~\ref{mech:poison} is about an excursion
  \emph{trap}, and a seed that never has the excursion never falls in.
  \item \textbf{Shaping} (\texttt{shape0}): success drops to 2/8 with
  three full never-serve collapses.  The honest nuance: shaping is not
  binary --- occasional seeds succeed on the sparse cost --- but it
  roughly triples the success probability.
  \item \textbf{Temperature annealing} (\texttt{anneal0}): the most
  instructive cell.  Failures are mostly graceful (ratios
  $1.28$--$2.46$), and on four of the five failed seeds the
  \emph{stochastic} readout is nearly optimal while the argmax is far
  off, e.g.\ (argmax vs stochastic, $\times\Lambda$):
  $94.2$ vs $42.0$;\; $52.5$ vs $41.7$;\; $314.0$ vs $43.1$;\;
  $51.7$ vs $41.3$.
  Without annealing, PPO \emph{learns} the right behavior but the
  argmax cannot \emph{extract} it --- the re-presentation forgiveness
  gap of Section~\ref{sec:forgiveness}, measured directly.
\end{itemize}

\subsection{A first action-space probe: the skip-all action}
\label{sec:skipall}

As a low-cost probe toward the reformulation of
Section~\ref{sec:blindness}, we added an optional third action:
\emph{skip-all} declines every remaining candidate in the current
action queue in one decision, making whole-tick idling a single,
readable choice instead of a chain of skips.  By construction skip-all
is value-degenerate with the skip chain (verified: a hand-coded
always-skip-all policy and an always-skip policy evaluate identically to
four decimals), so RVI benchmarks are unaffected; the policy head simply
gains a third logit, initialized pessimistically (bias $-3$,
${\approx}2\%$ initial probability) so that uniform initialization does
not grant whole-tick idling a $1/3$ prior.

The 8-seed result is negative but mechanistically clean: 3/8 seeds
within 5\% (vs 6/8 for the control), with four seeds collapsing --- and
on the collapsed seeds the stochastic readout remains moderate
($\times\Lambda$ between $51$ and $92$) while the argmax is
catastrophic ($352$--$372$).  Skip-all does not break \emph{learning};
it breaks \emph{extraction}: a wrong-sign near-tie on the skip-all
logit, harmless under the stochastic policy, is repeated
deterministically at every tick under the argmax and idles the system.
This sharpens the design requirement for the full reformulation: it is
not enough to make idling expressible in one action --- skipping must be
a \emph{commitment} (no re-presentation within the tick) so that
training itself feels the cost of wrong idling decisions, rather than
deferring the problem to extraction.

\subsection{Summary}

\begin{itemize}
  \item PPO, correctly adapted to the semi-MDP structure and stabilized
  by four ingredients (correct elapsed-period discounting,
  potential-based urgency shaping, value-target normalization, staggered
  resets) plus guarded temperature annealing for extraction, learns
  RVI-optimal policies on both benchmark instances.
  \item Experiment 3 (routing) is solved: 7--8/8 seeds within 5\% of
  RVI.  Experiment 2 (strategic idling) is solved per problem instance:
  7/8 seeds within 6\% under the tuned recipe, and 2--3 seeds with
  evaluation-based selection always suffice.
  \item Each ingredient is substantiated by a controlled 8-seed
  ablation; removing any one collapses success to 0--3/8, with failures
  landing on two exactly reproducible attractors (never-serve, ratio
  $9.876$; FIFO clone, ratio $1.738$).
  \item The residual seed lottery on Experiment 2 traces to
  advantage-weighted blindness at behaviorally flat decision boundaries,
  aggravated by re-presentation forgiveness at extraction time.  The
  skip-all probe shows that naive action-space patches inherit the
  extraction problem; the per-event reformulation, which removes
  re-presentation by construction, is the principled next step.
\end{itemize}

% ============================================================
\section{The queue-lateness objective}
\label{sec:ql}
% ============================================================

Everything so far optimized the binary deadline cost.  This section
repeats the program for the \emph{queue-lateness} (QL) objective ---
arguably the more natural cost when the depth of lateness matters ---
and finds an inverted failure landscape: the extraction machinery of
Section~\ref{sec:anneal} becomes largely unnecessary, but the policy
gradient itself drowns in the objective's heavy-tailed returns.  The
recipe that works is materially different, and the benchmark itself
required repair along the way.

\subsection{The reward function}
\label{sec:qlreward}

Queue-lateness charges, at every tick, the \emph{current excess} of every
waiting job (\texttt{reward\_type}~1):
\begin{equation}
  c_{\mathrm{QL}}(s) \;=\; \sum_n c_n \Bigl[
    \underbrace{\textstyle\sum_{t \,\in\, \mathrm{tracked}(n)} \max(0,\, t - d_n)}_{\text{tracked positions}}
    \;+\;
    \underbrace{\tfrac{\lambda_n}{\nu}\,\tfrac{(B_n-1)\,B_n}{2}\,\mathbf{1}\{B_n > 0\}}_{\text{expected tail}}
  \Bigr],
  \qquad B_n = t^{\mathrm{last}}_n - d_n,
  \label{eq:qlcost}
\end{equation}
where $t^{\mathrm{last}}_n$ is the deepest tracked age (the FIL itself at
tracking depth 1).  The tail term is exact in expectation: under
FIFO-within-type service, the jobs behind the FIL arrived as a Poisson
stream (density $\lambda_n/\nu$ per tick-slot) during the FIL's waiting
window, so a job that arrived $k$ ticks after the FIL has excess
$B_n - k$, and summing the arithmetic series gives \eqref{eq:qlcost}.
Because the cost is linear in per-job excesses, substituting the
conditional expectation is unbiased.

Two structural properties matter for everything that follows.  First,
\emph{QL is pre-deadline flat but post-deadline dense}: like the binary
cost it gives no signal before $d_n$, but past the deadline the per-tick
charge grows linearly with depth and sums over the whole queue --- deep
backlogs are punished superlinearly.  Second, a dimensional caveat found
during this work: with the parameter scalings of
Section~\ref{sec:setting}, QL cost levels are \emph{not} tick-rate
invariant (they scale $\approx$ linearly with $\nu$; measured
FIFO$\times\Lambda$ = 1781 / 3035 / 5266 at $\nu = 1.5/3/6$, versus the
binary cost's flat 74/71/70).  All results below are at $\nu = 3$, where
the objective is internally consistent; cross-tick-rate comparisons under
QL measure the cost definition rather than the system, until the extra
$1/\nu$ normalization is adopted.

\paragraph{Shaped variant.}
By direct analogy with Section~\ref{sec:shaping}, \texttt{reward\_type}~3
adds the same FIL urgency potential
$\Phi_n(t) = c_n \min(t, d_n)/d_n$: a constant accrual $c_n/d_n$ per
pre-deadline tick --- spreading exactly one \emph{first-late-tick} QL
charge ($c_n$) over the waiting window --- refunded in full when the job
enters service.  Cost neutrality was verified to four decimals (FIFO
evaluates to $722.5200$ per period under both rtype 1 and 3).

\subsection{Repairing the benchmark}
\label{sec:qlrvi}

The RVI reference under QL initially came out at
$\mathrm{RVI}\times\Lambda = 2492$ against
$\mathrm{FIFO}\times\Lambda = 3035$ (gap $1.22$; Experiment~3: $194.6$ vs
$152.0$, gap $1.28$, closely mirroring the binary gaps).  The repair was
forced by the learning results themselves: \emph{seven independently
trained policies evaluated below the computed optimum}, by up to 7\% ---
impossible if the reference were tight.  The cause is the truncation
loop's between-level criterion: the FIL clamp $M$ grows until $g^*$
changes by less than \texttt{rel\_tol} = 1\% between $M$ and $M{+}2$,
which is adequate for the binary cost (saturating in FIL) but stops too
early for QL, whose cost keeps growing with FIL depth through the tail
term in \eqref{eq:qlcost}.  Tightening to $10^{-3}$ moved the
Experiment~2 reference from $2492$ to $\mathbf{2350}$ ($-5.7\%$;
Experiment~3 moved only $0.2\%$ --- its reference was already sound).
A residual $\sim$1--2\% looseness remains (the best learned policies sit
at $0.98$--$0.99$ of the repaired reference), so within-$x$\% claims
below carry that uncertainty.

\subsection{What breaks: heavy-tailed returns, not extraction}
\label{sec:qlfailure}

Running the binary-cost recipe of Table~\ref{tab:hyper} on QL fails
completely --- and diagnostically.  All 8 seeds (and all 8 shaped-variant
seeds, and every ablation variant) converge to a \emph{bit-exact FIFO
clone} ($3034.5839$); Experiment~3 under QL solves normally (2/4 seeds
within 1\%), so the machinery is intact.  Three trace observations locate
the mechanism:

\begin{enumerate}
  \item The \emph{stochastic} readout is also FIFO ($\approx 3100$):
  unlike every binary-cost failure, this is a \textbf{learning} failure,
  not an extraction failure.  Consistently, the argmax-vs-stochastic gap
  that drove Section~\ref{sec:anneal} is essentially absent under QL ---
  as predicted from the cleaner threshold structure of the QL optimal
  policy.
  \item Per-rollout reward rates swing by an order of magnitude
  ($-380$ to $-7500$ per period) within healthy runs: QL returns are
  dominated by rare deep-lateness excursions whose cost is quadratic in
  depth via the tail term.  The 22\% FIFO-to-optimal improvement is
  buried under this variance at the $16 \times 512$ batch size.
  \item The temperature guard \emph{never engages} ($T = 1.0$ at the
  final update): its health signal inherits the same heavy tail, so the
  ``not degrading'' certificate is never issued (see
  \S\ref{sec:qlguard}).  Shaping is irrelevant here --- the shaped runs
  reproduce the unshaped results exactly, because the problem is noise,
  not signal density.
\end{enumerate}

In short, the failure landscape \emph{inverts} relative to the binary
cost: QL fixes the extraction problem for free and replaces it with a
variance problem.

\subsection{The QL recipe: average-reward objective at large batch}
\label{sec:qlrecipe}

The variance diagnosis points to the regime of Dai and Gluzman:
average-cost PPO with very large batches, built for queueing networks
with exactly this cost character.  Our implementation already contained
the average-reward mode (differential rewards $r - \hat\rho\,\Delta$,
$\gamma = 1$, with $\hat\rho$ a running estimate of the per-period
average reward; the critic then learns an RVI-style relative value
function).  A $2 \times 2$ scout --- \{discounted, average-reward\}
$\times$ \{$16 \times 512$, $128 \times 2048$\} --- separates the two
factors, and an 8-seed confirmation at 500 and 1000 updates completes the
picture:

\begin{itemize}
  \item \textbf{Small batch}: both objectives fail (average-reward small
  batch is the worst cell, confirming the earlier finding that the
  differential mode needs batch size).
  \item \textbf{Large batch, discounted}: partially works at 250 updates
  (best seed $0.96\times$ reference) but \emph{destabilizes with more
  training}: at 500 updates, 3/8 seeds collapse to never-serve and a
  previously good seed regresses to FIFO.  The discount horizon and the
  heavy tail interact badly once sharpening begins.
  \item \textbf{Large batch, average-reward}: the robust cell.  6/8
  seeds beat FIFO at 500 updates with a single collapse; at 1000 updates
  the sharpened seeds cluster at 2365--2627.  The differential objective
  measures every action against the current average cost rate, so a
  drift toward idleness is charged immediately --- there is no discount
  horizon for the tail catastrophe to hide behind.
\end{itemize}

At 1000 updates the bottleneck moved: \emph{every} seed whose behavior
temperature reached $T \approx 0.1$ landed in the good cluster, and every
seed stuck at $T = 1$ did not (4/8 engaged).  The remaining problem was
the guard itself.

\subsection{Guarding under heavy tails}
\label{sec:qlguard}

The guard of Section~\ref{sec:anneal} certifies health as
$\mathrm{EMA} \ge \mathrm{ref} - 0.05\,|\mathrm{ref}|$, where the
reference \emph{ratchets} up to the best EMA seen.  Under QL two
mechanisms defeat it.  \emph{Ratchet poisoning}: one lucky quiet stretch
(a few rollouts without deep-lateness excursions) sets an unrepeatable
reference; ordinary fluctuation then reads as permanent degradation, and
$T$ is pinned at 1 for the rest of training.  \emph{Non-arrival}:
descending from $T{=}1$ to $0.1$ requires $\approx 22$ net certified
steps; with the certificate failing half the time, the reflected random
walk rarely arrives.  Three independent one-line fixes were A/B-tested on
the four stock-guard-failing seeds:

\begin{itemize}
  \item \textbf{Noise-adaptive tolerance} ($\mathrm{tol} = 2\hat\sigma$
  of the health signal): engages 4/4 --- but sharpened one genuinely
  broken policy into a catastrophic deterministic one.  An
  always-permissive guard is blind annealing again; rejected on that
  ground despite two excellent rescues.
  \item \textbf{Robust health signal} (median over the 128 environments'
  per-period rates): engages 3/4 and, importantly, \emph{correctly
  refuses} to sharpen the broken policy.
  \item \textbf{Leaky ratchet} (reference relaxes 5\% toward the current
  EMA on each failed check): same qualitative profile, including the
  correct refusal.  The separation of speeds is the point: temperature
  backs off immediately on degradation while the reference forgets only
  slowly, so genuine decline is still caught.
\end{itemize}

The combined \textbf{robust + leaky} guard was confirmed on all 8 seeds
(Table~\ref{tab:qlfinal}): engagement rises from 4/8 (stock) to 7/8, the
$T{=}1$-stuck pathology disappears, and the one refusal is a seed whose
policy genuinely dies mid-training under every guard variant --- a
separate, unresolved failure mode that best-of-$k$ seed selection covers
in practice.

\begin{table}[h]
\centering
\caption{Final QL recipe on Experiment 2 (average-reward, $128 \times
2048$ batch, 1000 updates, robust+leaky guard), argmax
$\mathrm{cost}\times\Lambda$ vs repaired RVI $= 2350$, FIFO $= 3034$.}
\label{tab:qlfinal}
\begin{tabular}{lcccccccc}
\toprule
Seed & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\
\midrule
$\times\Lambda$ & 2645 & 2588 & 2663 & \textbf{2359} & 2544 &
\textbf{2333} & $\times$ & 2848 \\
ratio & 1.13 & 1.10 & 1.13 & \textbf{1.00} & 1.08 & \textbf{0.99} &
--- & 1.21 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Summary of the QL chapter}

\begin{itemize}
  \item QL inverts the binary failure landscape: extraction is free
  (argmax $\approx$ stochastic throughout), but heavy-tailed returns
  defeat small-batch policy gradients and fixed-tolerance guards; the
  binary recipe converges to a bit-exact FIFO clone on 8/8 seeds.
  \item The working recipe is \emph{average-reward + $64\times$ batch +
  tail-robust guarded annealing}: 7/8 seeds beat FIFO, best seeds at the
  (repaired) optimum, best-of-3 selection reliably $\approx$ optimal.
  The two chapters are symmetric: each objective needs the stabilizer
  matched to its own failure mode --- shaping against sparsity (binary),
  batch and robust statistics against tails (QL).
  \item Methodological yields: learned policies beating the benchmark
  exposed a 6\% RVI truncation bias (repaired via a tighter
  truncation-growth tolerance); the shaped QL variant is exactly
  cost-neutral but empirically unnecessary; QL cost levels are not
  tick-rate invariant under the current parameter scaling (flagged for
  any cross-tick study).
  \item Open: one seed exhibits mid-training policy death that no guard
  variant rescues (guards correctly refuse to sharpen it); its mechanism
  is distinct from all five binary-cost mechanisms and remains to be
  diagnosed.
\end{itemize}

% ============================================================
\section{Reformulating the action space: the per-event decomposition}
\label{sec:pe}
% ============================================================

The preceding chapters treated the action space as given and patched its
pathologies.  This section tests the diagnosis constructively: if
re-presentation forgiveness and credit dilution are really consequences
of the sequential candidate-queue decomposition, then a reformulated
action space should dissolve them --- and most of the stabilization
scaffolding with them.  It does, almost completely.

\subsection{Design}

In the \emph{per-event} mode, the decision epoch presents one entry per
\emph{idle server capacity unit} (pool-major order).  Each entry is a
single $(n{+}1)$-way decision: action $0$ idles this unit, action
$a \in \{1..n\}$ assigns type $a{-}1$'s FIL to it.  Infeasible types are
strictly masked (no dummy actions: value-degenerate action pairs would
pad DCL's rollout comparisons with noise-decided tie labels --- the
skip-all lesson).  A serve consumes its unit and the epoch \emph{resumes}
at the next unit after the FIL refresh; earlier idle decisions stay
committed until the next event.  Decision epochs occur at the same points
as in the candidate-queue mode, so the two spaces admit the same
physical policies: this was verified by three equivalence gates (FIFO
behavior and never-serve cost reproduce \emph{bit-exactly}; the RVI
optimum matches within solver tolerance on both instances).  What
changes is purely the \emph{credit structure}: a serve-vs-idle
counterfactual is one logit comparison carrying the full event-level
advantage, instead of a sign smeared across a skip chain and forgiven by
re-presentation.  The per-decision action space is $O(n)$ regardless of
server count --- the scaling motivation.

\subsection{Results: the pathologies dissolve}

Table~\ref{tab:pe} shows the full battery on Experiment~2 (8 seeds,
unchanged hyperparameters from each objective's tuned recipe).

\begin{table}[h]
\centering
\caption{Per-event action space, Experiment 2, argmax
$\mathrm{cost}\times\Lambda$.  Binary refs: RVI $=40.78$, FIFO $=70.38$;
QL refs: RVI $=2350$, FIFO $=3034$.  ``anneal'' = guarded temperature
annealing of Section~\ref{sec:anneal}.}
\label{tab:pe}
\begin{tabular}{lcccccccc}
\toprule
Seed & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\
\midrule
PPO binary, anneal on & 95.4 & \textbf{40.78} & \textbf{40.78} & 40.87 &
89.4 & \textbf{40.78} & 98.0 & 96.3 \\
PPO binary, \textbf{anneal off} & \textbf{40.98} & \textbf{40.78} &
\textbf{40.78} & \textbf{40.78} & \textbf{40.78} & \textbf{40.78} &
\textbf{40.78} & \textbf{40.78} \\
DCL binary (FIFO base) & 41.1 & \textbf{40.78} & 41.0 & \textbf{40.78} &
\textbf{40.78} & \textbf{40.78} & 264 & 44.4 \\
DCL queue-lateness & 2511 & 2769 & 2556 & 2748 & 2766 & 2769 & 2697 &
2769 \\
\bottomrule
\end{tabular}
\end{table}

Three results, each contradicting nothing and confirming everything the
mechanism analysis predicted:

\begin{enumerate}
  \item \textbf{PPO without annealing: 8/8 at the exact RVI optimum.}
  The Experiment-2 seed lottery --- the central residual problem of the
  binary chapter --- is gone, with \emph{less} machinery than the
  candidate-queue space ever needed (recall: the old space without
  annealing succeeded on 3/8).  The argmax--stochastic relation is
  healthy everywhere (argmax $\le$ stochastic): nothing is left for
  extraction to break.
  \item \textbf{The scaffolding turns harmful.}  With annealing ON, 4/8
  seeds fail with the classic extraction signature (argmax $\gg$
  stochastic).  In the old space, sharpened rollouts were needed to
  train wrong-sign near-ties; in the new space there are no wrong signs
  left to fix, and mid-training sharpening only narrows the data
  distribution.  A stabilizer is not objective- or space-neutral: it is
  matched to a failure mode, and outlives its usefulness with it.
  \item \textbf{DCL escapes its fixed points.}  From a plain
  deterministic FIFO base --- no stochastic-base widening needed ---
  6/8 seeds reach the optimum under the binary cost (candidate-queue
  reference: 3/8, with five exact FIFO clones), because the
  $(n{+}1)$-way labels natively express the serve-this-type-vs-idle
  counterfactuals that the ordering-and-label designs of the old space
  tried to manufacture.  Under queue-lateness, 8/8 seeds beat FIFO with
  \emph{zero} catastrophic collapses (reference: 2/8 good, 3/8
  collapsed), at 1.07--1.18$\times$ the repaired optimum.
\end{enumerate}

Caveats, honestly: one DCL-binary seed produced a new failure mode
(264, worse than never studied before in DCL-binary --- under
investigation); DCL-QL quality trails the best PPO-QL seeds; and the
queue-lateness PPO cell in the per-event space (which needs the
large-batch average-reward recipe --- QL's variance problem is
orthogonal to the action space) is in flight at the time of writing,
together with an 8-seed Experiment-3 confirmation and a first scaling
scout on the RVI-intractable 6-type/5-server instance.

\subsection{Interpretation}

The candidate-queue chapters and this one form a single argument.  The
five failure mechanisms were diagnosed as \emph{mismatches between the
RL abstraction and the decision structure}, and each fix was a patch
holding that mismatch at bay: shaping against sparsity, guards against
blind sharpening, readout batteries against extraction, seed selection
against the lottery.  The per-event reformulation removes the mismatch
at its source, and the patches become unnecessary --- or actively
counterproductive --- exactly as the mechanism story predicts.  For
scaling, the reformulated space is the natural primary formulation: its
per-decision action set is independent of server count, its credit
assignment is event-local, and it needs a strictly smaller recipe.

\end{document}
